Creating Documents with Rmd or Rnw
======================

You'll use substantially the same structure to write a document using
markdown or $\LaTeX$, that is, with a `.Rmd` or `.Rnw` file type.

##  Why would you want to use one versus the other?
* $\LaTeX$ compiles nicely to PDF, Rmd to HTML
* Rmd is easier to write and compile, but offers less control over formatting
* If you already know $\LaTeX$, Rnw isn't really any harder.

## The Boiler Plate

Files will typically start with an R-chunk that sets things up the way you want it.
(This can be hidden from view using `include=FALSE` inside the chunk tag.)
```{r message=FALSE, tidy=FALSE}
require(mosaic)                         # load packages that will be used
require(knitr)                          # so you can run chunks

opts_chunk$set(                         # set some defaults for R output
  fig.width=8,                          # generate plots 8 inches wide
  fig.height=3,                         # generate plots 3 inches tall
  out.width="80%",                      # width of figures in HTML
  fig.align="center",                   # center figures
  dev = 'svg',                          # create svg graphics
  tidy = FALSE                          # don't tidy up the R code
  )                                     # there are many other options, too

trellis.par.set(theme=col.mosaic())     # change a bunch of color settings
trellis.par.set(fontsize=list(text=15)) # set font size for plots
```

## Some examples


Let's take a look at the function 
$$f(x) = sin(3x)$$
and its derivative and antiderivative.
The style of the "chunk" differs, but the R commands are the same in Rmd and Rnw.

```{r}
f <- makeFun( sin(3*x) ~ x )
g <- D( f(x) ~ x )
F <- antiD( f(x) ~ x )
plotFun( f(x) ~ x, x.lim=c(0,4), ylim=c(-pi,pi) )
plotFun( F(x) ~ x, add=TRUE, col=2)
plotFun( g(x) ~ x, add=TRUE, col=3)
```   


