Modeling Tides
========================================================

```{r echo=FALSE,results="hide",label="options"}
opts_chunk$set(fig.width=6,fig.height=3,out.width="3in")
```

The file ```hawaii.csv``` contains hour-by-hour measurements of water level in Pearl Harbor, Hawaii, over approximately 4 days.

```{r warning=FALSE,error=FALSE,message=FALSE}
library(mosaic,quietly=TRUE)
trellis.par.set(theme=col.mosaic())
tides = fetchData("hawaii.csv")
xyplot(water~time, data=tides)
```

How to model these data?  They are repetitive, which suggests a sine-wave model.  We need to know the period, amplitude, phase, and offset.  Looking at the graph suggests a period of about 24 hours, as seems sensible.  

```{r}
f0 = fitModel( water ~ A + B*sin(2*pi*(time-T0)/P), 
               data=tides, start=list(P=24))
xyplot(water~time, data=tides, pch=20)
plotFun(f0(time)~time, add=TRUE, lwd=3, col="red")
```
Here, ```T0``` is the phase of the sine function.

Another way to address the phase issue is to add a cosine of the same period.
Fitting that sort of model, and comparing to the time-shift parameterization shows that they are effectively the same.
```{r}
f1 = fitModel( water ~ A + B*sin(2*pi*time/P)+C*cos(2*pi*time/P), 
               data=tides, start=list(P=24))
xyplot(water~time, data=tides, pch=20)
plotFun(f1(time)~time, add=TRUE, lwd=4)
plotFun(f0(time)~time, add=TRUE, col="red")
```

According to the model, what's the period?

That's a pretty rough model. Perhaps that's because there are often two tides per day.  So there's something going on with a period of about 12 hours.  A new model:
```{r}
f2 = fitModel( water ~ A + B*sin(2*pi*time/P1)+C*cos(2*pi*time/P1) +
               D*sin(2*pi*time/P2)+E*cos(2*pi*time/P2), 
               data=tides,
               start=list(P1=24,P2=12))
xyplot(water~time, data=tides,pch=20)
plotFun(f2(time)~time, add=TRUE,lwd=3)
```

Much better!  

## Questions
1. What's the relationship between the two periods?
2. Do you get a much better model with three or more periods?  What should those new periods be?
3. The model is quite good, yet the fundamental period is not 24 hours.  Why not?

## Resources
* A source for minute-by-minute data from [NOAA](http://tidesandcurrents.noaa.gov/tsunami/#)
* A bit of [nomenclature](http://oceanservice.noaa.gov/education/kits/tides/media/supp_tide07a.html)
